Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2022-03-24 fls-lightning-talk/slides.Rmd
minor tweak
updated slides per discussion at meeting

1. Define a generative model for graph: 
$G \mid z_1, ..., z_n, \vec{\theta} \sim P(\vec{z}, \vec{\theta})$.
2. Develop a method for obtaining estimators: 
$f(G) = (\hat{\vec{z}}, \hat{\vec{\theta}})$.
3. Identify asymptotic properties of estimators: 
$(\hat{\vec{z}}, \hat{\vec{\theta}}) \to (\vec{z}, \vec{\theta})$.

## Overview

1. Block Models and the Popularity Adjusted Block Model

2. Generalized Random Dot Product Graphs

3. Connecting the PABM to the GRDPG

4. Community Detection for the PABM

# Block Models

@@

Erdos-Renyi Model (1959)

* $P_{ij} = \theta$ (not a block model)
* 1 parameter $\theta$

Stochastic Block Model (Lorrain and White, 1971)

* $P_{ij} = \theta_{z_i z_j}$
* $K (K + 1) / 2$ parameters $\theta_{kl}$

Degree Corrected Block Model (Karrer and Newman, 2011)

* $P_{ij} = \theta_{z_i z_j} \omega_i \omega_j$
* $K (K + 1) / 2 + n$ parameters $\theta_{kl}$, $\omega_i$

Popularity Adjusted Block Model (Sengupta and Chen, 2017)

* $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
* $K n$ parameters $\lambda_{ik}$

@@

Let each vertex $i \in [n]$ have $K$ popularity parameters
$\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
Then $A \sim \PABM(P)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$,  
e.g., if $z_i = k$ and $z_j = l$, $P_{ij} = \lambda_{il} \lambda_{jk}$.

@@

**Notation**: $A \sim \PABM(\{\lambda^{(kl)}\}_K)$.

# Generalized Random Dot Product Graphs

Random Dot Product Graph $A \sim \RDPG(X)$  
(Young and Scheinerman, 2007)

* Latent vectors $x_1, ..., x_n \in \mathbb{R}^d$ such that 
$x_i^\top x_j \in [0, 1]$
* $A \sim \BG(X X^\top)$ where 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$

Generalized Random Dot Product Graph $A \sim \GRDPG_{p, q}(X)$  
(Rubin-Delanchy, Cape, Tang, Priebe, 2020)

* Latent vectors $x_1, ..., x_n \in \mathbb{R}^{p+q}$ such that 
$x_i^\top I_{p, q} x_j \in [0, 1]$ and $I_{p, q} = \blockdiag(I_p, -I_q)$
* $A \sim \BG(X I_{p, q} X^\top)$ where 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$

If latent vectors $X_1, ..., X_n \iid F$, then we write 
$(A, X) \sim \RDPG(F, n)$ or $(A, X) \sim \GRDPG_{p, q}(F, n)$.

@@

To embed in $\mathbb{R}^d$,

1. Compute $A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$ 
where $\hat{\Lambda} \in \mathbb{R}^{d \times d}$ and 
$\hat{V} \in \mathbb{R}^{n \times d}$.  
For RDPG, use $d$ greatest eigenvalues; 
for GRDPG, use $p$ most positive and $q$ most negative eigenvalues.

2. For RDPG, let $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$; 
for GRDPG, let $\hat{X} = \hat{V} |\hat{\Lambda}|^{1/2}$.

@@

RDPG: $\max\limits_i \|\hat{X}_i - W_n X_i \| \stackrel{a.s.}{\to} 0$
(Athreya et al., 2018)  
GRDPG: 
$\max\limits_i \|\hat{X}_i - Q_n X_i \| \stackrel{a.s.}{\to} 0$
(Rubin-Delanchy et al., 2020)

# Connecting the PABM to the GRDPG

@@

```{r, out.width = '50%', fig.height = 2.5, fig.width = 2.5}
par(mar = rep(1.75, 4))
plot(X, asp = 1, col = z, xlab = NA, ylab = NA, main = 'SBM: Point masses')
```

```{r, out.width = '50%', fig.height = 2.5, fig.width = 2.5}
par(mar = rep(1.75, 4))
omega <- rbeta(n, 1, 1)
dc.matrix <- omega %*% t(omega)
P.dcbm <- P * dc.matrix
dcbm.eigen <- eigen(P.dcbm)
X.dcbm <- dcbm.eigen$vectors[, 1:2] %*% diag(dcbm.eigen$values[1:2] ** .5)
plot(X.dcbm, asp = 1, col = z, xlab = NA, ylab = NA, main = 'DCBM: Rays')

```

@@

$$X = \begin{bmatrix}
\Lambda^{(1)} & \cdots & 0 \\
0 & \ddots & 0 \\
0 & \cdots & \Lambda^{(K)}
\end{bmatrix}$$

\vspace*{2\baselineskip}

$$U \in \mathbb{O}(K^2)$$

@@

**Remark 1** (orthogonality of subspaces):
If $y_i^\top$ and $y_j^\top$ are two rows of $X U$ corresponding 
to different communities, then $y_i^\top y_j = 0$.

**Remark 2** (non-uniqueness of the latent configuration):  
If $A \sim \GRDPG_{p, q}(Y)$, then $A \sim \GRDPG_{p, q}(Y Q)$ for any $Q$ in 
the indefinite orthogonal group with signature $p, q$.

**Remark 3**: Communities correspond to subspaces even with linear transformation $Q \in \mathbb{O}(p, q)$, but this may break the orthogonality property.

# Community Detection for the PABM

@@

## Sparse Subspace Clustering

**Corollary**: The ASE of $A \sim \PABM(\{\lambda^{(kl)}\}_K)$ lies near a collection of $K$-dimensional subspaces in $K^2$ dimensions.

**Algorithm**: Sparse Subspace Clustering (Elhamifar & Vidal, 2009):

1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
subject to $x_i = X^\top c$ and $c^{(i)} = 0$.  
This is typically performed via LASSO: 
$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i}^\top c\|_2^2 + \lambda \|c\|_1$
2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
3. Construct affinity matrix $B = |C| + |C^\top|$.

## Sparse Subspace Clustering

**Theorem** (KTT):

Let

* $P_n$ describe the edge probability matrix of the PABM with $n$ vertices, and 
$A_n \sim \BG(P_n)$;
* $\hat{V}_n$ be the matrix of eigenvectors of $A_n$ corresponding to the 
$K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues.

Then

* For some $\lambda > 0$ and $N < \infty$, $\sqrt{n} \hat{V}_n$ obeys the Subspace Detection Property with probability 1 when $n > N$.

Remarks:

* For large $n$, we can identify $\lambda$ for SDP (Wang and Xu, 2016).
* SDP does not guarantee community detection.

@@

@@

>>>>

1. Define a generative model for graph: 
$G \mid z_1, ..., z_n, \vec{\theta} \sim P(\vec{z}, \vec{\theta})$.
2. Develop a method for obtaining estimators: 
$f(G) = (\hat{\vec{z}}, \hat{\vec{\theta}})$.
3. Describe asymptotic properties of estimators: 
$(\hat{\vec{z}}, \hat{\vec{\theta}}) \to (\vec{z}, \vec{\theta})$.

@@

@@

Let each vertex $i \in [n]$ have $K$ popularity parameters
$\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
Then $A \sim \PABM(\{\lambda_{ik}\}_K)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$,  
e.g., if $z_i = k$ and $z_j = l$, $P_{ij} = \lambda_{il} \lambda_{jk}$.

@@

## Generalized Random Dot Product Graph

**Def** Generalized Random Dot Product Graph  
(Rubin-Delanchy, Cape, Tang, Priebe, 2020)

$A \sim \GRDPG_{p, q}(X)$ iff

* Latent vectors $x_1, ..., x_n \in \mathbb{R}^{p+q}$ such that 
$x_i^\top I_{p, q} x_j \in [0, 1]$ and $I_{p, q} = \blockdiag(I_p, -I_q)$
* $A \sim \BG(X I_{p, q} X^\top)$ where 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
  * $P(\text{edge between } v_i, v_j) = x_i^\top I_{p,q} x_j$

@@

To embed in $\mathbb{R}^{p+q}$,

1. Compute $A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$ 
where $\hat{\Lambda} \in \mathbb{R}^{(p+q) \times (p+q)}$ and 
$\hat{V} \in \mathbb{R}^{n \times (p+q)}$ 
by using the $p$ most positive and $q$ most negative eigenvalues.

2. Let $\hat{X} = \hat{V} |\hat{\Lambda}|^{1/2}$.

@@

$\max\limits_i \|\hat{X}_i - Q_n X_i \| = O_P \Big( \frac{(\log n)^c}{n^{1/2}} \Big)$
(Rubin-Delanchy et al., 2020)

@@

```{r, out.width = '100%', fig.height = 4, fig.width = 4}
# par(mar = rep(1.75, 4))
plot(X, asp = 1, col = z, xlab = NA, ylab = NA, main = 'SBM: Point masses')
```

@@

@@

@@

@@

Simulation setup:

1. $z_1, ..., z_n \iid \Cat(1/K, ..., 1/K)$
2. $\lambda_{ik} \iid \Betadist(a_{ik}, b_{ik})$  
$a_{ik} = \begin{cases} 2 & z_i = k \\ 1 & z_i \neq k \end{cases}$ 
$b_{ik} = \begin{cases} 1 & z_i = k \\ 2 & z_i \neq k \end{cases}$
3. $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$
4. $A \sim \BG(P)$

@@

## Conclusion

1. The PABM is a recently developed flexible block model that can be used to describe graphs with community structure.

2. The GRDPG, which can describe all block models, motivates a spectral approach to statistical inference on graphs.

3. Under the GRDPG framework, the PABM with $K$ communities can be induced by a latent configuration in $\mathbb{R}^{K^2}$ consisting of $K$ $K$-dimensional subspaces that are orthogonal to each other.

4. The latent configuration of the PABM under the GRDPG framework leads to an intuitive method for community detection with nice theoretical asymptotic properties.
